---
title: "Patient Data Exploration"
author: "Mark Dunning"
date: "21 April 2016"
output: html_document
---

```{r}
library(dplyr)
library(stringr)

patients <- read.delim("patient-data.txt")
patients <- tbl_df(patients)
patients_clean <- mutate(patients, Sex = factor(str_trim(Sex)))
patients_clean <- mutate(patients_clean, Height= as.numeric(str_replace_all(patients_clean$Height,pattern = "cm","")))
```

- We want to calculate the Body Mass Index (BMI) for each of our patients
- $BMI = (Weight) / (Height^2)$ 
    + where Weight is measured in Kilograms, and Height in Metres
- Create a new BMI variable in the dataset
- A BMI of 25 is considered overweight, calculate a new variable to indicate which individuals are overweight
- For a follow-on study, we are interested in overweight smokers
    + clean the `Smokes` column to contain just `TRUE` or `FALSE` values
- How many candidates (Overweight and Smoker) do you have?
- (EXTRA) What other problems can you find in the data?


```{r}
##YOUR CODE HERE.....

```

